---
title: "User Guide: PLSandO"
author: "Maider Aguerralde Martin"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This guide provides a comprehensive walkthrough of the PLSandO tool, designed to handle standard Principal Component Analysis (PCA), Partial Least Squares (PLS), PLS Discriminant Analysis (PLS-DA), Multi-block structures (MBPCA, MBPLS and MBPLSDA), and multilevel cases. Developed by the BiostatOmics Lab, it is publicly available on the [BiostatOmics GitHub repository](https://github.com/BiostatOmics/PLSandO/). 

Beyond model fitting, PLSandO incorporates a wide range of functionalities for data preprocessing, result exploration and interpretation, variable selection, and outlier detection, enabling a complete start-to-end analysis workflow from raw data to model evaluation and interpretation.

# Installation

Currently, the package can be directly installed from GitHub using the `devtools` R package by:

```{r, eval=FALSE}
install.packages("devtools")
devtools::install_github("BiostatOmics/PLSandO")
```

# Load PLSandO

```{r, echo=FALSE,include=FALSE}
devtools::load_all(".")
```


```{r, echo=FALSE,eval=FALSE}
library(PLSandO)
```

# PCA

We will use the classic `iris` dataset to demonstrate the basic workflow of PCA with PLSandO: Preprocessing -> Model Fitting -> Visualization.

## Data Preparation

PLSandO includes the `Preparing()` function, which handles the transformation of categorical variables for their inclusion in the different dimensionality reduction techniques available in PLSandO. In addition, it allows filtering of variables with near-zero variance and the exclusion of variables or observations with a high proportion of missing values.

Although `Preparing()` can be applied independently, it is internally integrated into the PCA workflow using conservative parameter settings aimed at preventing numerical or computational issues during model fitting. The internal preprocessing applies only minimal and extreme filters — such as the removal of variables with zero variance or the removal of variables with almost no available values — that would otherwise prevent the PCA algorithm from being properly computed.

In the following example, categorical variables are encoded for inclusion in multivariate models, variables with (near-)zero variance are removed, and variables or observations with a high proportion of missing values are excluded. The filtering thresholds shown here are intentionally conservative and are mainly intended to avoid numerical issues during PCA computation rather than to perform aggressive data cleaning.

As output, `Preparing()` also returns a concise summary describing the preprocessing steps applied and the modifications performed on the original dataset, allowing users to easily track how the data have been transformed.

```{r}
data("iris")

# Preprocess the data
x_prepared = Preparing(iris,includeFactors = TRUE, CVfilter = 0.00001, excludeNA = 0.2)

#Observe the modifies dataset
head(x_prepared$x)

#Observe the function modifications
x_prepared$removed
```
The output of `x_prepared$x` could be used to its introduction into `pca()`function. However, this conservative thresholds are already taken into account into so they would be already considered when running the `pca()`main function.

## Running PCA

A PCA model can easily be run by:

```{r}
mypca = pca(iris)
```
## Results Visualization

### Essential Visualization

The `pcaPlot()` function is a versatile wrapper for different plots for PCA results visualization.

First the code for visualizing a Scree Plot is introduced:

```{r}
# 1. Scree Plot: Check explained variance
pcaPlot(mypca, type = 'scree')
```

Then, the codes for the Score and Loading plots are introduced. In the Score plot, the observations will be colored by the Species variable including the 95\% confidence ellipses. In the Loading plot, variables will be colored by their contribution (so the importance of the variable in the component). In both cases, first and second components will be shown.

```{r}

# 2. Score Plot: Observations colored by Species
pcaPlot(mypca, type = 'scores', colBy = 'Species', labels = FALSE)

# 3. Loadings Plot: Variable contributions
pcaPlot(mypca, type = 'loadings', shape = 'arrow', colBy = 'contrib', labels = TRUE)
```

Additionally, PLSandO includes the option to create biPlot which overlays the Score and Loading plots. It allows us to see the relationship between samples and variables simultaneously. For instance, you can identify which variables are responsible for a specific cluster of samples.

```{r}
# 4. Biplot: Observations and variables in the same space
pcaPlot(mypca, type = 'biplot', colBy = 'Sepal.Length', labels = c(FALSE, TRUE))
```

The Correlation Plot projects variables onto a unit circle. The proximity of a variable to the circle's edge indicates how well it is represented by the current components. Arrows pointing in the same direction indicate high positive correlation.

```{r}
# 5. Correlation Plot: Evaluates variable representation and correlations
pcaPlot(mypca, type = 'corr', shape = 'arrow', colBy = 'cos2', labels = TRUE)
```

Finally, The R2 Plot provides a bar-chart summary of how much variance of each individual variable is captured by the model in each component. 

```{r}
# 6. R2 Plot: Explained variance per individual variable in each component
pcaPlot(mypca, type = 'R2')
```

### Working with new Observations

One of the core strengths of PLSandO is the ability to project new data onto an existing PCA space, even when the new observations contain missing values (NAs). This allows users to assess whether observations not included in the original PCA model are consistent with the same underlying population and whether they can be meaningfully combined with the original data for further analysis.

```{r}
# Exploratory set (125 obs) and New observations (25 obs). We will not ocnsider Species variable in this example
x_train = iris[1:125, -5]
x_new = iris[126:150, -5]

# Introduce some NAs in the new data
x_new[1, 1] = NA
x_new[3, 4] = NA

# Fit model with training data
pca_model = pca(x_train)

# Project new observations (shown as distinct points)
pcaPlot(pca_model, type = 'scores', newObs = x_new)
pcaPlot(pca_model, type = 'scores', newObs = x_new, colBy = 'Sepal.Length')
```

## Running Multi-block PCA (MBPCA)

When the data is organized into logical blocks (e.g., measurements from different omics, sensors, ...), MBPCA allows for a consensus analysis.

As supported by the literature (e.g., Westerhuis et al.), performing a Multi-block PCA is mathematically equivalent to performing a standard PCA on a **concatenated dataset**, provided that the blocks have been properly **block-scaled**. 

Block-scaling is essential because it prevents blocks with a larger number of variables from dominating the model. Our tool automates this process through the `scaling` parameter, ensuring that each block contributes equally to the global consensus (the "Super-scores").

Multi-block data can be handled in two ways using this tool. In both cases, data must be provided in a list:

### Option A: Concatenation with Block-Scaling

The most widely used approach in the literature consists of concatenating the datasets and applying hard-block scaling. This ensures that the variance of each block is normalized relative to its size, allowing for a fair integration of data sources.

This option can easily be applied by:

```{r}
# Create blocks: Petal measurements vs Sepal measurements
X_blocks = list('Petal' = iris[,c(3,4)], 'Sepal' = iris[,c(1,2)])

# Applying 'softBlock' scaling on the concatenated matrix
pca_hard = pca(X_blocks, ncomp = 2, scaling = 'hardBlock')
```
Scores are shown to show the users the equivalence between the two approaches.

```{r}
head(pca_hard$scores)
```

In this case, all previously introduced plots can be used.

### Option B: Using MBPCA

Even though the results are equivalent to the concatenated approach, providing the data as a list and using `algo = 'mbpca'` offers two advantages:

1. **Computational Efficiency:** In cases of **really high-dimensional data** (e.g., Omics data or high-resolution spectroscopy), handling individual datasets in a list is often faster and more memory-efficient. R can manage these smaller chunks more easily than a single, massive concatenated matrix.
2. **Block-level Diagnostics:** Using the MBPCA algorithm allows the tool to calculate the specific **contribution of each block** to the global explained variance (R^2). This is essential for understanding which data source is driving the model.

This is the recommended path for high-dimensional data or when block diagnostics are required.

```{r}
# Fit Multi-block PCA
mb_model = pca(X_blocks, algo = 'mbpca')

head(mb_model$Sscores)
```

All previously presented plots are also allowed in MBPCA, but results are show at block and super level. An example will be showed with the visualization of the Scores.

```{r, fig.show='hold', out.width='50%'}
# Score Plot for each block and the super level. Observations colores by the Petal.Width
pcaPlot(mb_model, type = 'scree')
pcaPlot(mb_model, type = 'scores', colBy = 'Petal.Width')
```

We can also look at the explained variance of the variables and the blocks to the model.

```{r, fig.show='hold', out.width='50%'}
# R2 
pcaPlot(mb_model, type = 'R2')
```

# PLS

The `pls()` function allows modeling the relationship between a predictor matrix `X` and a response matrix `Y`. This is particularly useful when predictors are highly collinear or high-dimensional.

In line with PCA, the `Preparing()` function can be called prior to fitting a PLS model to handle essential matrix preprocessing. However, if the data is passed directly to the `pls()` function, the model automatically applies conservative default parameters to ensure the data is appropriately preprocessed for PLS modelling. 

## Running Single-response PLS (PLS1)

Even if the dataset we have been using is not intended for this end, we will continue using the iris dataset to show case how to apply `pls()` function of `PLSandO` package. We will just take one of the variables as the response variable and we will try to model its relation with the rest of the variables.

A PLS model can easily be run by:

```{r}
# Define X and Y
X = iris[, 1:3, drop = FALSE]
Y = iris[, 4, drop = FALSE]

# Fit PLS model
mypls = pls(X, Y)
```

## Running Multi-response PLS (PLS2)

If Y contains multiple response variables, the package automatically performs PLS2. A PLS2 can be easily run by:

```{r}
# Define X and Y
X2 = iris[, 1:2, drop = FALSE]
Y2 = iris[, 3:4, drop = FALSE]

# Fit PLS model
mypls2 = pls(X2, Y2)
```

## Results exploration

Once the model has been fitted, some brief summary can be explored to evaluate the performance of the model has been fitted. The `pls()` output includes information about the explained variance, the predictive ability through evaluated via cross-validation, and some other metrics that allowed to determine the optimal number of components and which can be used to assess the reliability of the model's predictions.

These results can easily be evaluated via:

```{r}
mypls$summary
```

Results from the cross-validation can also be explored via:

```{r, eval = FALSE}
mypls$cv_results
```

## Results visualization

The `plsPlot()` function is a versatile wrapper for different plots for PLS results visualization.

First the code for visualizing a R2vsQ2 Plot is introduced:

```{r}
# 1. R2vsQ2
plsPlot(mypls, type = 'R2vsQ2')
```

Then, the codes for the Score, Loading and Weight plots are introduced. In the Score plot, the observations will be colored by the Species variable including the 95\% confidence ellipses. In the Loading plot, variables will be colored according to if they relate to `X` or `Y` matrices. In both cases, first and second components will be shown. In the Weights plot, no additional functionalities will be asked.

```{r}
# 2. Score Plot: Observations colored by Species
plsPlot(mypls, type = 'scoresX', colBy = as.data.frame(iris[, 5, drop = FALSE]), ellipses = TRUE)

# 3. Loading Plot
plsPlot(mypls, type = 'loadings')

# 4. Weight Plot
plsPlot(mypls, type = 'weights')

```

Additionally, PLSandO includes the option to create some plots to validate the PLS model created. 

```{r}
# 5. Validating the internal relation between u and t scores
plsPlot(mypls, type = 'linearity')

# 6. Validating there are no overfitting problems
plsPlot(mypls, type = 'overfitting')
```

We can also analyze the explained variance of each variable per component via:

```{r}
#7. R2 per variable and component
plsPlot(mypls, type = 'R2')
```

## Prediction

As a supervised model, once the PLS model has been fitted, we can predict values for new observations using the `plsPredict()` function. Beyond generating numeric predictions, the function itself can create a plot to project these new observations into the latent space. This allows you to verify that the new data points belong to the same population as the training set, ensuring that the predictions for those observations are reliable and not based on extreme extrapolations.

```{r}
# Splitting data for prediction
X_train = iris[1:125, 1:3, drop = FALSE]
Y_train = iris[1:125, 4, drop = FALSE]
X_new = iris[126:150, 1:3, drop = FALSE]

# Re-fitting and predicting
mypls_train = pls(X_train, Y_train)
pred = plsPredict(mypls_train, new = X_new)
head(pred)
```

We can also create extra plots also projecting new observations via:

```{r}
# Biplot with new observations
plsPlot(mypls_train, type = 'biplot', newObs = X_new)
```

## Missing values

The package is robust to missing data, as PLS naturally handles NAs during training and prediction. Beyond this, the model could use the latent space structure to provide reliable imputations for missing values.

```{r}
# Introducing missing values
X[1,1] = NA
Xnew = iris[126:150,1:3,drop=FALSE]
Xnew[2,1] = NA

mypls_na = pls(X,Y)

#Return X matrix without missing data using the latent structure
X_complete = imputeNipals(mypls_na, scaled = F)
head(X_complete)
#Predict values even there are missing values
head(plsPredict(mypls_na,Xnew))
```

## Outlier detection

PLSandO also provides functionalities to identify influential outliers.

```{r, fig.show='hold', out.width='50%'}
# Outlier detection and contribution analysis
out = plsOutliers(mypls)
plsOutlierContrib(mypls, outliers = out)
```

## Variable selection

For variable selection, the `plsVarSel()` function identifies the most influential predictors in your model, allowing you to simplify your model and focus on the key variables that drive the model. It supports the variable selection according to different criteria: by Variable Importance in Projection (VIP), by the significant Multivariate Correlation (sMC), p-values obtained from Jack-Kniffing or by Permutation sampling. 

```{r}
# Summary of models variables and Jack-Knife resampling results
mypls$coefficients_summary

# Identify key predictors using Permutation testing
plsVarSel(mypls, type = 'Perm')

# Identify key predictors using VIP scores
plsVarSel(mypls, type = 'VIP')
```

## Running Multi-block PLS (MBPLS)

When your predictors are organized into logical groups — such as different omics layers, distinct experimental conditions, or separate sensor arrays — Multi-block PLS (MBPLS) allows you to model their relationship with a response `Y` while preserving the underlying block structure. As with MBPCA, the package automates permits to apply block-scaling to apply the PLS on the concatenated matrices or to apply the MBPLS modelling.

To run an MBPLS, provide you can provide `X` data as a named list where each element is a block, and specify `algo = 'mbpls'`.

```{r}
# Organize data into blocks (e.g., Petal vs Sepal)
X_blocks = list('Petal' = iris[, c(3, 4)], 'Sepal' = iris[, c(1, 2)])
Y_target = iris[, 3, drop = FALSE]

# Fit Multi-block PLS
mymbpls = pls(X_blocks, Y_target, algo = 'mbpls')
```

MBPLS allows for exploring the Block Importance on Projection (BIP).

```{r}
mymbpls$bip
```

All previously presented plots are also allowed in MBPLS, but results are show at block and super level. An example will be showed with the visualization of the Scores and loadings.

```{r, fig.show='hold', out.width='50%'}
# Global Score Plot (Super-scores) colored by Species
plsPlot(mymbpls, type = 'scoresX', colBy = as.data.frame(iris[, 5, drop=FALSE]), ellipses = TRUE)

# Block Loadings to see variable contributions within their respective blocks
plsPlot(mymbpls, type = 'loadingsX')
```

# PLSDA

## Running PLS Discriminant Analysis (PLS-DA)

When the response variable `Y` is categorical (e.g., treatment groups, phenotypes, or disease status), PLS-DA is the appropriate approach to model the relationship between the predictor matrix `X`and the response `Y`. In PLS-DA, latent components maximize class separation while capturing the covariance structure between `X` and `Y`.

In line with PCA, the `Preparing()` function can be called prior to fitting a PLS-DA model to handle essential matrix preprocessing. However, if the data is passed directly to the `plsda()` function, the model automatically applies conservative default parameters to ensure the data is appropriately preprocessed for PLS modelling. 

In PLSandO, the `plsda()` function can easily be applied by:

```{r}

X = iris[, 1:4]
Y = iris[, 5, drop = FALSE]

# Fit PLS-DA model
myplsda = plsda(X, Y)
```

## Results exploration

Once the model has been fitted, similarly to `pls` case, some brief summary can be explored to evaluate the performance of the model has been fitted. The `pls()` output includes information about the explained variance, the predictive ability through evaluated via cross-validation, and some other metrics that allowed to determine the optimal number of components and which can be used to assess the reliability of the model's predictions.

These results can easily be evaluated via:

```{r}
myplsda$summary
```

## Results visualization

The `plsdaPlot()` function serves as a versatile wrapper for visualizing PLS-DA results. It inherits most of the plotting capabilities found in `plsPlot()`, providing a consistent interface for model exploration.

In PLS-DA an additional plotting option `type='ncomp'` is introduced. While the package internally determines the optimal number of components based on the Balanced Error Rate (BER) to ensure robust performance across different class sizes, this visualization allows for a more comprehensive assessment. By using this plot, users can evaluate the choice of components through multiple lenses, including various model performance indicators and classification error metrics, to ensure the most stable model selection.

```{r, fig.show='hold', out.width='50%'}
# Performance and Model selection
plsdaPlot(myplsda, type = 'R2vsQ2')
plsdaPlot(myplsda, type = 'ncomp')
```

To better visualize group differentiation, `plsdaPlot()` automatically incorporates 95\% confidence ellipses for the `Y` categories. These ellipses provide a statistical boundary that helps in assessing the separation and overlap between different classes in the score space.

```{r, fig.show='hold', out.width='50%'}
# Space visualization
plsdaPlot(myplsda, type = 'scoresX')
plsdaPlot(myplsda, type = 'biplot')
```

```{r, fig.show='hold', out.width='50%'}
# Variables contribution
plsdaPlot(myplsda, type = 'loadings')
plsdaPlot(myplsda, type = 'weights')
```

## Prediction

Prediction can easily be performed by `plsdaPredict()` function which outputs the categorical prediction of the model and the prediction of the pls model.

```{r}
X_train = iris[1:125, 1:4]
Y_train = iris[1:125, 5, drop = FALSE]

X_test = iris[126:150, 1:4]

# Fit and Predict
myplsda_subset = plsda(X_train, Y_train)
predictions = plsdaPredict(myplsda_subset, new = X_test)

head(predictions$prediction)
head(predictions$plsprediction)
```

## Missing values

It is also robust against missing values.

```{r}
# Example with missing data and new observations
X_train = iris[1:125, 1:4]
Y_train = iris[1:125, 5, drop = FALSE]
X_train[1, 1] = NA # Introducing a missing value

X_test = iris[126:150, 1:4]
X_test[2, 1] = NA  # Missing value in new data

# Fit 
myplsda_subset = plsda(X_train, Y_train)

# Visualize new observations in the existing space
plsdaPlot(myplsda_subset, type = 'biplot', newObs = X_test)
```

## Variable selection

Variable selection is also available for the same methodologies introduced in `plsVarSel()`.

```{r}
# Variable selection via Jackknife
plsdaVarSel(myplsda, type = 'Jack')
```

## Outliers

Finally, some additional analysis can be perfomed for outlier detection.

```{r, fig.show='hold', out.width='50%'}
# Outlier detection and their contribution to the model
out = plsdaOutliers(myplsda)
plsdaOutlierContrib(myplsda, outliers = out)
```

## Running Multi-block PLS Discriminant Analysis (MBPLSDA)

Similarly to MBPLS, the MBPLS-DA algorithm is used when predictors are structured in blocks but the goal is classification. 

To run an MBPLS-DA, provide the `X` data as a list and set `algo = 'mbplsda'`.

```{r}
# Organize data into blocks
X= list('Petal' = iris[, c(3, 4)], 'Sepal' = iris[, c(1, 2)])
Y = iris[, 5, drop = FALSE]

# Fit Multi-block PLS-DA
mymbplsda = plsda(X, Y, algo = 'mbplsda')
```

All previously presented plots are also allowed in MBPLSDA, but results are show at block and super level. An example will be showed with the visualization of the Scores and loadings.

```{r, fig.show='hold', out.width='50%'}
# Score plot with ellipses for classes
plsdaPlot(mymbplsda, type = 'scoresX')

# Evaluate linearity and overfitting
plsdaPlot(mymbplsda, type = 'overfitting')
```

Predicting with multi-block data requires the new input to maintain the same list structure as the training blocks.

```{r, fig.show='hold', out.width='33%'}
# Define training and test sets for blocks
X_train = list('Petal' = iris[1:125, c(3, 4)], 'Sepal' = iris[1:125, c(1, 2)])
X_test = list('Petal' = iris[126:150, c(3, 4)], 'Sepal' = iris[126:150, c(1, 2)])
Y_train = iris[1:125, 5, drop = FALSE]

# Model and Predict
mymbplsda2 = plsda(X_train, Y_train, algo = 'mbplsda', ncomp = 2)
mb_preds = plsdaPredict(mymbplsda2, new = X_test)
```

# Multilevel Analysis

The `multilevel()` function is specifically designed for study designs with repeated measures or crossover structures. It accounts for variability across different levels of the data (for example, between subjects or clusters) so that these sources of variation are properly separated from the effects of interest.

The multilevel approach handles this by decomposing the data matrix into three separate parts before applying multivariate models:
- Offset (`X_m`): Captures the mean of all variables in `X`.
- Within-subject variation (`X_w`): Captures the changes occurring within the same individual across different conditions. This is typically the variation of interest.
- Between-subject variation (`X_b`): Captures the intrinsic differences between individuals.

To demonstrate the functionality of the `multilevel()` function, we will use a repeated-measures dataset from the mixOmics package.

## Running multilevel PCA

```{r,warning=FALSE}
# Loading mixOmics data and preparing the data as needed for PLSandO
library(mixOmics)
data(vac18)
X = vac18$gene
design = vac18$sample
names(design) = rownames(X)
head(design)#show the design format
```

Once the data is prepared the multilevel will be applied to account for the study design when exploring the data.

```{r}
# Fit a Multilevel PCA
multipca = multilevel(X, design = design)
names(multipca)
```

### Results Visualization

The `multilevelPlot()` function automatically accounts for the structure of multilevel objects, and all options available in `pcaPlot()` can also be used here. For exploratory analysis, both within- and between-subject variations can be visualized jointly. In addition, the `$Within` and `$Between` objects of a multilevel object are themselves PCA objects, meaning that all standard PCA functionalities can be applied to them separately. Some examples are shown below.

```{r}
#Create the scree plot
multilevelPlot(multipca, type ='scree' )

#Obtain the same plot for the Within part using pcaPlot
pcaPlot(multipca$Within, type ='scree')
```

To have a look at the scores, we can simply run:

```{r}
#Explore the score plot
multilevelPlot(multipca, type='scores')
```

## Running multilevel PLSDA

The dataset also contains class information that can be used for supervised analysis. We will use it to illustrate how to run a PLS-DA model with repeated measurements.

```{r}
# Loading mixOmics data and preparing the data as needed for PLSandO
Y = as.data.frame(vac18$stimulation)
rownames(Y) = rownames(X)
```

Once the data are prepared, the multilevel approach can be applied to account for the study design during model fitting and exploration.

```{r}
# Fit a Multilevel PLSDA
multiplsda = multilevel(X, Y, design = design, method = 'plsda')
names(multiplsda)
```

### Results Visualization

The `multilevelPlot()` function automatically handles the structure of multilevel objects, regardless of the method used. When `method = 'plsda'`, it focuses on visualizing within-subject variation.

First, the optimal number of components should be determined. For this purpose, we examine the plot with `type = 'ncomp'`.

```{r}
#Create the scree plot based on different metrics
multilevelPlot(multiplsda, type ='ncomp' )
```

Most classification error–based metrics indicate that 3 components are optimal. The model is therefore re-fitted using this number of components.

```{r}
multiplsda = multilevel(X, Y, design = design, method = 'plsda', ncomp = 3)
```

Finally, the results can be explored.

```{r}
#Explore the model results
multiplsda$model$summary
```

```{r}
#Create the score plot
multilevelPlot(multiplsda, type='scoresX')
```

```{r}
#Look at the model coefficients
multilevelPlot(multiplsda, type = 'coef')
```

### Prediction

Prediction in a multilevel context requires both the new data (`new`) and the information specifying the subject to which each sample belongs (`design`). Prediction can easily be performed by `multilevelPredict()` function which outputs the categorical prediction of the model and the prediction of the pls model.

In the following example, a subset of samples is held from the training data and used to evaluate the prediction procedure.

```{r}
#Prepare the training data
X = vac18$genes[-c(4,17,26,37,7,28,39),]
Y = as.data.frame(vac18$stimulation[-c(4,17,26,37,7,28,39)])
rownames(Y) = rownames(X)

design = vac18$sample[-c(4,17,26,37,7,28,39)]
names(design) = rownames(Y)

#Fit the model
mymultiplsda = multilevel(X, Y, design, method = 'plsda')

#Select the optimal number of components
multilevelPlot(mymultiplsda, type ='ncomp')
```

```{r}
#Fit the model with the optimal number of components
mymultiplsda = multilevel(X, Y, design, method = 'plsda', ncomp = 3)
```

```{r}
#Test data preparation
Xtest = vac18$genes[c(4,17,26,37,7,28,39),]
designTest = vac18$sample[c(4,17,26,37,7,28,39)]
names(designTest) = rownames(Xtest)

#Test data prediction
myprediction = multilevelPredict(mymultiplsda, new = Xtest, design = designTest)
head(myprediction$prediction)
head(myprediction$plsprediction)
```

## Running multilevel PLS

The PLS model is run in a similar way. In this case, we do not include a full example with results, but the following code illustrates how the model can be fitted.

```{r, eval = FALSE}
multipls = multilevel(X, Y, design = design, method = 'pls')
```








